---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

We are working with the data of the number of steps taken by an anonymous individual over a period of two months, collected with the help of a personl activity monitoring device. The device collects data at 5 minute intervals throughout the day.

## Loading and preprocessing the data
The data is given to us in the form of a zip file. We import this data by first unzipping this file to obtain a .csv file named *activity.csv* which is then read in using the `read.csv()` command in R.
```{r reading_data}
unzip('activity.zip')
data<-read.csv('activity.csv')
file.remove('activity.csv')
```
Here we can see that once the data is read in, the `file.remove()` function is used to remove the unzipped file to preserve the original state of the repository.  
```{r checking_format}
str(data)
head(data)
```
The data has been read in as a data frame with **17568 rows** and **3 columns**. As this is a widely used format in R, we will proceed to answer the remaining questions without making any further changes to this data.


## What is mean total number of steps taken per day?
To gain some insight to the data, let us first plot a **histogram** of the number of steps taken each day.
```{r histogram1}
par(mfrow = c(1,1))
sdata<-tapply(data$steps,data$date,sum,na.rm=TRUE)
total.steps.per.day<-sdata
hist(total.steps.per.day, xlab = "Total Steps per Day")
```  


When reading using the `read.csv()` function, all strings are automatically read in as factors, thus we can directly use the `tapply()` function to find the sum of the number of steps taken in each interval for each day. Here we also make use of the `na.rm = TRUE` option to avoid any errors due to missing data.  
Let us calculate the mean and the median of the *Total number of steps taken per day* across the 2 month period.
```{r mean_median1}
mean_data<-mean(sdata)
print(paste('Mean:',as.character(mean_data)))
median_data<-median(sdata)
print(paste('Median:',as.character(median_data)))
```

## What is the average daily activity pattern?
Each day is divided into 288 intervals of 5 mins each, ranging from 0 to 2355. Let us visualise the average number of steps taken in each interval across the period of 2 months.
```{r interval1}
mean_across_interval<-tapply(data$steps,data$interval,mean,na.rm=TRUE)
x_val<-as.numeric(names(mean_across_interval))
plot(x_val,mean_across_interval,type="l", xlab = "Interval",ylab = "Steps")
```  


From this plot it is easy to see that the maximum of this value occours between the intervals 500 and 1000. Let us verify this...
```{r maxinterval}
max_value<-which.max(mean_across_interval)
mean_across_interval[max_value]
```
Thus, we see that the maximum number of steps if usually taken in the interval **835-840** with a mean value of 206.


## Imputing missing values
It is common to find that the data provided for analysis has many missing values. Here too, this is the case. 
```{r missing_data}
unzip('activity.zip')
data<-read.csv('activity.csv')
file.remove('activity.csv')
sum(!complete.cases(data))
```
Here we see that there are 2304 rows which have missing entries and this may have an impact on our analysis. To fill up this missing data, we will use a simple scheme where we input the mean steps taken in that interval in all intervals where there is no data provided.
```{r fill_missing_data}
data1<-data

for(i in 1:17568)
{
  if(is.na(data1$steps[i]))
  {
    data1$steps[i]<-mean_across_interval[as.character(data1$interval[i])]
  }
}
```
Here, we have created a new dataset `data1` which has all the missing values filled in.
Let us now compare the results of the analysis between the two datasets.
```{r comparison, fig.height=10}
sdata1<-tapply(data1$steps,data1$date,sum)
total.steps.per.day.after.correction<-sdata1

par(mfrow = c(2,1))
hist(total.steps.per.day, xlab = "Total Steps per Day", ylim = c(0,40))
hist(total.steps.per.day.after.correction, xlab = "Total steps per day", ylim = c(0,40))

mean_data1<-mean(sdata1)
print(paste('Old Mean:',as.character(mean_data)))
print(paste('New Mean:',as.character(mean_data1)))
median_data1<-median(sdata1)
print(paste('Old Median:',as.character(median_data)))
print(paste('New Median:',as.character(median_data1)))
```
Thus we can clearly see in both the histogram as well as the mean and median that the new data now has many observations where the total steps taken in a day moved from 0 to the daily mean, which is 10766.

## Are there differences in activity patterns between weekdays and weekends?
To compare the walking trends between weekdays and weekends, we make use of the new dataset that we have created.
```{r weekdays_weekends, fig.height=10}
data1$day_of_week<-weekdays(as.Date(data1$date))
data1$weekend<- data1$day_of_week=="Saturday" | data1$day_of_week =="Sunday"
tempList<-split(data1,factor(data1$weekend))
df1<-tempList$'FALSE'
df2<-tempList$'TRUE'

head(data1)

par(mfrow = c(2,1))

mean_across_interval_df1<-tapply(df1$steps,df1$interval,mean,na.rm=TRUE)
x_val_df1<-as.numeric(names(mean_across_interval_df1))
plot(x_val_df1,mean_across_interval_df1,type="l", xlab = "interval", ylab = "mean steps", ylim = c(0,250))
title(main = "Mean steps across each interval on Weekdays")

mean_across_interval_df2<-tapply(df2$steps,df2$interval,mean,na.rm=TRUE)
x_val_df2<-as.numeric(names(mean_across_interval_df2))
plot(x_val_df2,mean_across_interval_df2,type="l", xlab = "interval", ylab = "mean steps", ylim = c(0,250))
title(main = "Mean steps across each interval on Weekends")
```  


Here we add 2 new columns to the dataset, which signify the day of the week and also signify if the day is a weekend or not. We plot a time series graph to understand the difference between the trends.
